---
title: "Cardiometabolic Differences and Changes"
author: "J. O'Connor"
date: "2025-09-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Cardiometabolic Differences and Changes

This markdown completed the analysis comparing cardiometabolic markers across groups and looking at longitudinal changes 

### Data Loading

We load all the data of interest

```{r data_load, echo=FALSE}
#All Metadata objects have been stored in one location
metadata_objects <- readRDS("~/Zimbabwe-Cardiometabolic-Analysis/Data/R_Data/metadata_objects.RDS")

  #base: all processed metadata (including 16S and 18S add ons)
  base <- metadata_objects$base

#We filter to include only baseline values
  base_t1 <- base %>%
              filter(Week==0)

  

```

### Normality Asssesment 

Clinical variables are selected and tested for normality. 

```{r Normality, echo=FALSE, message=FALSE, warning=FALSE}
#All base_t1line variables of interest are pulled out 
#Categorical Variables
myvars_cat <- c("Location","Gender","BlastocystisPresent_rare")

#Continuous variables
myvars_cont <- c("Visit_Age", "BMI", "HDL_mg-dL", "LDL_mg-dL", "Trig_mg-dL", "Glucose_mg-dL", "IL-6_pg_mL", "CRP_mg_L", "HIV-1_viral_load", "Faith_Diversity", "Shannon_Diversity","Cotri_Length_Month","CD4+_count")

myvars_all <- c(myvars_cat,myvars_cont)


#Normality is assessed 
##Blank vectors are generated and filled with normal and not normal variables 
myvars_cont_norm <- c()
myvars_cont_non_norm <- c()

#Below Shapiro's tests are used for normality of all continuous variables
for (j in 1:length(myvars_cont)) {
  #Variable selected and data pulled 
  var_j <- myvars_cont[j]
  var_j_col <- which(colnames(base_t1)==var_j)
  base_t1[,var_j_col] <- as.numeric(unlist(base_t1[,var_j_col]))
  
  #Normality is assessed by a Shapiro test
  shap.j <- shapiro.test(as.numeric(unlist(base_t1[,var_j_col])))
  shap.j.p <- shap.j$p.value 
  
  #If the p value is under 0.05 it is not normal otherwise it is normal 
  if (shap.j.p <0.05) {
    myvars_cont_non_norm <- append(myvars_cont_non_norm, var_j)
  } else {
    myvars_cont_norm <- append(myvars_cont_norm, var_j)
  }
}


#Below we get an indication of which are normal and which are not 

if (length(myvars_cont_norm)>0) {
  if (length(myvars_cont_norm)>1) {
    and_var <- myvars_cont_norm[length(myvars_cont_norm)]
    base_t1_vars <- setdiff(myvars_cont_norm, and_var)
    base_t1_vars_str <- paste(base_t1_vars, collapse =", ")
    and_var_str <- paste0(", and", and_var)
    print(paste0("Variables ", base_t1_vars_str, and_var_str, " are normally distributed")) 
  } else {
    print(paste0(myvars_cont_norm, " is normally distributed")) 
  }
}

if (length(myvars_cont_non_norm)>0) {
  if (length(myvars_cont_non_norm)>1) {
    and_var <- myvars_cont_non_norm[length(myvars_cont_non_norm)]
    base_t1_vars <- setdiff(myvars_cont_non_norm, and_var)
    base_t1_vars_str <- paste(base_t1_vars, collapse =", ")
    and_var_str <- paste0(", and ", and_var)
    print(paste0("Variables ", base_t1_vars_str, and_var_str, " are NOT normally distributed")) 
  } else {
    print(paste0(myvars_cont_non_norm, " is NOT normally distributed")) 
  }
}


```

###Data Group Quantification and Table Generation

base_t1line data is pulled and tables are generated with statistics reflecting normality 

```{r Quantification and Table Generation, echo=FALSE, message=FALSE, warning=FALSE}


#base_t1line tables are generated 

#Main Group
main_tab_1_pre <- CreateTableOne(vars=myvars_all, strata="Cohort", data=base_t1, argsNonNormal = c(myvars_cont_non_norm), argsNormal = c(myvars_cont_norm), includeNA=FALSE)



##Additional pairwise comparisons
base_t1.gp12 <- base_t1[which(as.character(base_t1$Cohort) %in% c("Naïve","Exp")),]
base_t1.gp13 <- base_t1[which(as.character(base_t1$Cohort) %in% c("Naïve","HC")),]
base_t1.gp23 <- base_t1[which(as.character(base_t1$Cohort) %in% c("Exp", "HC")),]

##Additional pairwise tables are made for downstream analysis

#Allvariables are put into one group c
main_tab_1.gp12_pre <- CreateTableOne(vars=myvars_all, strata="Cohort", data=base_t1.gp12, argsNonNormal = myvars_cont_non_norm, argsNormal = myvars_cont_norm)
main_tab_1.gp13_pre <- CreateTableOne(vars=myvars_all, strata="Cohort", data=base_t1.gp13, argsNonNormal = myvars_cont_non_norm, argsNormal = myvars_cont_norm)
main_tab_1.gp23_pre <- CreateTableOne(vars=myvars_all, strata="Cohort", data=base_t1.gp23, argsNonNormal = myvars_cont_non_norm, argsNormal = myvars_cont_norm)

#For our visualization we are going to change the mean (SD)s to meadian (IQRs). To do this the print command is used and specify the all contonuus variables are not normal (even though Shannon is)

main_tab_1.gp12 <- as.data.frame(print(main_tab_1.gp12_pre, nonnormal=myvars_cont))
main_tab_1.gp13 <- as.data.frame(print(main_tab_1.gp13_pre, nonnormal=myvars_cont))
main_tab_1.gp23 <- as.data.frame(print(main_tab_1.gp23_pre, nonnormal=myvars_cont))

main_tab_1 <- as.data.frame(print(main_tab_1_pre, nonnormal=myvars_cont))
```


###Post HOC testing

P-values are adjusted for multiple comparisons

```{r Post HOC, echo=FALSE, message=FALSE, warning=FALSE}



#We set viral load to zero for our healthy controls
base_t1$`HIV-1_viral_load`[which(as.character(base_t1$Cohort)=="HC")] <- 0

#We generate a blank data frame that we add to which will include IQRs of the groups then adjusted p-values
cont_table <- data.frame("Variable"=c(),
                         "Naïve"=c(),
                         "Exp"=c(),
                         "HC"=c(),
                         "Naïve_Exp"=c(),
                         "Naïve_HC"=c(),
                         "Exp_HC"=c() 
                         )

#Now we go through each variable and do the post hoc testing
for (j in 1:length(myvars_cont)) {
  var_j <- myvars_cont[j]
  
  #We generate a simple temproray data file
  data_temp_j <- base_t1[which(!is.na(base_t1[,var_j])),which(colnames(base_t1) %in% c("Cohort",var_j))]
  
  #We name the continuous variable 
  colnames(data_temp_j)[which(colnames(data_temp_j)==var_j)] <- "Value"
  
  #Make sure it is numeric
  data_temp_j$Value <- as.numeric(data_temp_j$Value)
  
  
  #If the variable is not normal ...
  if (var_j %in% myvars_cont_non_norm) {
  #Dunn's test performed on Kruskall wallace using the dunn_test fuction
  stat_j <- dunn_test(data=data_temp_j,
                      Value ~Cohort,
                      p.adjust.method = "bonferroni")
  
  #We pull the pairwise p-values doing NA is there is no value for any pair
pg12 <- (function(x) if(length(x) == 0) NA else x)(stat_j$p.adj[which(stat_j$group1 %in% c("Naïve","Exp") & stat_j$group2 %in% c("Naïve","Exp"))])
pg13 <- (function(x) if(length(x) == 0) NA else x)(stat_j$p.adj[which(stat_j$group1 %in% c("Naïve","HC") & stat_j$group2 %in% c("Naïve","HC"))])
pg23 <- (function(x) if(length(x) == 0) NA else x)(stat_j$p.adj[which(stat_j$group1 %in% c("Exp","HC") & stat_j$group2 %in% c("Exp","HC"))])
  
  }
  
  #If the variable is normal...
  if (var_j %in% myvars_cont_norm) {
  
  # Parametric pairwise t-tests with Bonferroni adjustment
  stat_j <- pairwise.t.test(data_temp_j$Value, data_temp_j$Cohort, p.adjust.method = "bonferroni")

# To pull the p-values from this matrix:
pg12 <- (function(x) if(length(x) == 0 || is.na(x)) NA else x)(stat_j$p.value["Exp","Naïve"])
pg13 <- (function(x) if(length(x) == 0 || is.na(x)) NA else x)(stat_j$p.value["HC", "Naïve"])
pg23 <- (function(x) if(length(x) == 0 || is.na(x)) NA else x)(stat_j$p.value["HC", "Exp"])
  

  }
  
  #We pull the varaible rows from the Median IQR tables generated above
  
  var_row <- which(substr(rownames(main_tab_1),1,nchar(var_j))==var_j)
  
  #We add those median values to the conintuous table
  
  cont_table_row_j<- data.frame("Variable"=var_j,
                         "Naïve"=main_tab_1$Naïve[var_row],
                         "Exp"=main_tab_1$Exp[var_row],
                         "HC"=main_tab_1$HC[var_row],
                         "Naïve_Exp"=pg12,
                         "Naïve_HC"=pg13,
                         "Exp_HC"=pg23 
                         )
  
  cont_table <- rbind(cont_table,cont_table_row_j)

}

#The same is done for the categorical Variables

cat_table <- data.frame("Variable"=c(),
                         "Naïve"=c(),
                         "Exp"=c(),
                         "HC"=c(),
                         "Naïve_Exp"=c(),
                         "Naïve_HC"=c(),
                         "Exp_HC"=c() 
                         )

#We will pull out subsets and do chi-squares followed by p value adjustments
for (k in myvars_cat) {
 #We pull the combo names from the p adjusted table to match the columns
 comb_1 <- c("Exp","Naïve")
 comb_2 <- c("HC","Naïve")
 comb_3 <- c("Exp","HC")
 #We then generate subsets of data
 data_temp_k.gp_comb_1 <- base_t1[which(as.character(base_t1$Cohort) %in% comb_1 & !is.na(base_t1[,k])),which(colnames(base_t1) %in% c("Cohort",k))]
 data_temp_k.gp_comb_2 <- base_t1[which(as.character(base_t1$Cohort) %in% comb_2 & !is.na(base_t1[,k])),which(colnames(base_t1) %in% c("Cohort",k))]
 data_temp_k.gp_comb_3 <- base_t1[which(as.character(base_t1$Cohort) %in% comb_3 & !is.na(base_t1[,k])),which(colnames(base_t1) %in% c("Cohort",k))]
 
 #We nrename columns to generalize for the chiquare
 colnames(data_temp_k.gp_comb_1)[which(colnames(data_temp_k.gp_comb_1)==k)] <- "Value"
 colnames(data_temp_k.gp_comb_2)[which(colnames(data_temp_k.gp_comb_2)==k)] <- "Value"
 colnames(data_temp_k.gp_comb_3)[which(colnames(data_temp_k.gp_comb_3)==k)] <- "Value"
 
 #We perform the chi-square
 chi_k_comb_1 <- chisq.test(data_temp_k.gp_comb_1$Cohort,data_temp_k.gp_comb_1$Value)
 chi_k_comb_2 <- chisq.test(data_temp_k.gp_comb_2$Cohort,data_temp_k.gp_comb_2$Value)
 chi_k_comb_3 <- chisq.test(data_temp_k.gp_comb_3$Cohort,data_temp_k.gp_comb_3$Value)
 
 bon_adjust_k <- p.adjust(c(chi_k_comb_1$p.value,chi_k_comb_2$p.value,chi_k_comb_3$p.value), method="bonferroni")


   var_row <- which(substr(rownames(main_tab_1),1,nchar(k))==k)
     
  #We add those median values to the conintuous table
  
  cat_table_row_k<- data.frame("Variable"=k,
                         "Naïve"=main_tab_1$Naïve[var_row],
                         "Exp"=main_tab_1$Exp[var_row],
                         "HC"=main_tab_1$HC[var_row],
                         "Naïve_Exp"= bon_adjust_k[1],
                         "Naïve_HC"= bon_adjust_k[2],
                         "Exp_HC"=bon_adjust_k[3] 
                         )
  
  cat_table <- rbind(cat_table,cat_table_row_k)
 
}

```





###Final Table Generation

Now we ajoin the adjusted pairwise values to the main table

```{r Final Table, echo=FALSE, message=FALSE, warning=FALSE}


#We generate a first line with the number of samples 
n_gp1 <- nrow(base_t1[which(as.character(base_t1$Cohort)=="Naïve"),])
n_gp2 <- nrow(base_t1[which(as.character(base_t1$Cohort)=="Exp"),])
n_gp3 <- nrow(base_t1[which(as.character(base_t1$Cohort)=="HC"),]) 

comb_tab_1_pre <- as.data.frame(t(c("Sample Number", n_gp1, n_gp2, n_gp3," "," ", " ")))
colnames(comb_tab_1_pre)<- colnames(cat_table)

final_tab_1 <- rbind(comb_tab_1_pre,cat_table,cont_table)

write.csv(final_tab_1, file="../../Data/Metadata/Exported_Table1.csv")

```

