---
title: "Blastocystis Analysis"
author: "J. O'Connor"
date: "2025-09-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R markdown to perform all analysis comparison Blastocystsis to different markers

### Function Sourcing

First the functions are all sourced
```{r Functions, message=FALSE, warning=FALSE}
#To access the functions directory why must direct to the parent directory
setwd("../..")

#A list of the function files is pulled
function_files <- list.files(paste0(getwd(),"/Functions"), pattern = "\\.R$", full.names = TRUE)
#Those functions are then sourced
lapply(function_files, source)
```

###Data is Loaded
```{r cars}

#All 18S objects have been stored in one location
euk_18S_objects <- readRDS("~/Zimbabwe-Cardiometabolic-Analysis/Data/R_Data/euk_18S_objects.RDS")

#18S objects
  #base: all processed metadata
  base <- euk_18S_objects$base
  #Euk_18S_mdf: 18S mdf data frmae
  Euk_18S_mdf <- euk_18S_objects$Euk_18S_mdf
  #base_with_18S: metadata with with 18S added
  base_with_18S <- euk_18S_objects$base_with_18S
  #Zim_all_sg_genera_pre: Unfiltered shotgun taxonomic data
  base_with_18S <- euk_18S_objects$base_with_18S
  #fecal.tax.plot.data: 16S phyloseq object 
  fecal.tax.plot.data <- euk_18S_objects$fecal.tax.plot.data

#Previusly Determined confounders are loaded
conf_vars <- readRDS("~/Zimbabwe-Cardiometabolic-Analysis/Data/R_Data/conf_vars.RDS")

#Confounders
  #conf_vars_all: All confounders in one vector
  #conf_vars_HDL: Confounders of HDL
  #conf_vars_LDL: Confounders of LDL
  #conf_vars_Trig: Confounders of Trig
  #conf_vars_Glucose: Confounders of Glucose
  conf_vars_all <- conf_vars$conf_vars_all
  conf_vars_HDL <- conf_vars$conf_vars_HDL
  conf_vars_LDL <- conf_vars$conf_vars_LDL
  conf_vars_Trig <- conf_vars$conf_vars_Trig
  conf_vars_Glucose <- conf_vars$conf_vars_Glucose


```

###Statistics 


###Barplot Across Study Period
```{r}


Euk_18S_mdf$SampleID <- Euk_18S_mdf$Sample
Euk_18S_mdf$PID <- substr(Euk_18S_mdf$Sample,1,6)

Euk_18S_mdf$OtherGens <- other_generator(data=Euk_18S_mdf, 
                                         name_field="Genus", 
                                         type="Percentage", 
                                         threshold=.03, 
                                         abundance_field="Abundance", 
                                         id_var="PID")

Euk_18S_mdf <- inner_join(Euk_18S_mdf, base[,c(setdiff(colnames(base),colnames(Euk_18S_mdf)),"SampleID")], by="SampleID")

Euk_18S_mdf_t1 <- Euk_18S_mdf[which(Euk_18S_mdf$Week==0),]


Euk_Counts_t1 <- table(Euk_18S_mdf_t1$Genus)
Euk_Counts_Co_t1 <- table(Euk_18S_mdf_t1$Cohort, Euk_18S_mdf_t1$Genus)



Euk_18S_mdf$DisplayWeek <- paste0((as.numeric(substr(Euk_18S_mdf$Sample,8,8))-2)*24, " Weeks")




all_18S_IDs <- unique(Euk_18S_mdf$Sample)
Blast_IDs <- Euk_18S_mdf$Sample[which(Euk_18S_mdf$Genus=="Blastocystis" & Euk_18S_mdf$Abundance>0)]
Ent_IDs <- Euk_18S_mdf$Sample[which(Euk_18S_mdf$Genus=="Entamoeba" & Euk_18S_mdf$Abundance>0)]
Kom_IDs <- Euk_18S_mdf$Sample[which(Euk_18S_mdf$Genus=="Komagataella" & Euk_18S_mdf$Abundance>0)]
Str_IDs <- Euk_18S_mdf$Sample[which(Euk_18S_mdf$Genus=="Strigeidida" & Euk_18S_mdf$Abundance>0)]

Ent_Blast_data <- base

Ent_Blast_data$DisplayWeek <- paste0(Ent_Blast_data$Week, " Weeks")

Ent_Blast_data$Blastocystis <- ""
Ent_Blast_data$Entamoeba <- ""
Ent_Blast_data$Komagataella <- ""
Ent_Blast_data$Strigeidida <- ""

Ent_Blast_data$Blastocystis[which(Ent_Blast_data$SampleID %in% all_18S_IDs)] <- "No"
Ent_Blast_data$Entamoeba[which(Ent_Blast_data$SampleID %in% all_18S_IDs)] <- "No"
Ent_Blast_data$Komagataella[which(Ent_Blast_data$SampleID %in% all_18S_IDs)] <- "No"
Ent_Blast_data$Strigeidida[which(Ent_Blast_data$SampleID %in% all_18S_IDs)] <- "No"

Ent_Blast_data$Blastocystis[which(Ent_Blast_data$SampleID %in% Blast_IDs)] <- "Yes"
Ent_Blast_data$Entamoeba[which(Ent_Blast_data$SampleID %in% Ent_IDs)] <- "Yes"
Ent_Blast_data$Komagataella[which(Ent_Blast_data$SampleID %in% Kom_IDs)] <- "Yes"
Ent_Blast_data$Strigeidida[which(Ent_Blast_data$SampleID %in% Str_IDs)] <- "Yes"

Ent_Blast_data_long <- pivot_longer(data=Ent_Blast_data,
                                      cols=c("Blastocystis","Entamoeba","Komagataella","Strigeidida"),
                                      names_to = "Organism",
                                      values_to = "Present")

EukPresentData <- Ent_Blast_data_long[which(Ent_Blast_data_long$Present=="Yes"),]

EukPresentData_Count <- as.data.frame(table(EukPresentData$DisplayWeek,EukPresentData$Cohort, EukPresentData$Organism))

colnames(EukPresentData_Count) <- c("DisplayWeek","Cohort","Organism","Count")

EukPresentData_Count$Percent <- NA

EukPresentData_Count$Percent[which(EukPresentData_Count$Cohort=="Naïve" & EukPresentData_Count$DisplayWeek=="0 Weeks")] <- EukPresentData_Count$Count[which(EukPresentData_Count$Cohort=="Naïve" & EukPresentData_Count$DisplayWeek=="0 Weeks")]/65
EukPresentData_Count$Percent[which(EukPresentData_Count$Cohort=="Exp" & EukPresentData_Count$DisplayWeek=="0 Weeks")] <- EukPresentData_Count$Count[which(EukPresentData_Count$Cohort=="Exp" & EukPresentData_Count$DisplayWeek=="0 Weeks")]/33
EukPresentData_Count$Percent[which(EukPresentData_Count$Cohort=="HC" & EukPresentData_Count$DisplayWeek=="0 Weeks")] <- EukPresentData_Count$Count[which(EukPresentData_Count$Cohort=="HC" & EukPresentData_Count$DisplayWeek=="0 Weeks")]/40

EukPresentData_Count$Percent[which(EukPresentData_Count$Cohort=="Naïve" & EukPresentData_Count$DisplayWeek=="24 Weeks" )] <- EukPresentData_Count$Count[which(EukPresentData_Count$Cohort=="Naïve" & EukPresentData_Count$DisplayWeek=="24 Weeks")]/44
EukPresentData_Count$Percent[which(EukPresentData_Count$Cohort=="Exp" & EukPresentData_Count$DisplayWeek=="24 Weeks")] <- EukPresentData_Count$Count[which(EukPresentData_Count$Cohort=="Exp" & EukPresentData_Count$DisplayWeek=="24 Weeks")]/32
EukPresentData_Count$Percent[which(EukPresentData_Count$Cohort=="HC" & EukPresentData_Count$DisplayWeek=="24 Weeks")] <- EukPresentData_Count$Count[which(EukPresentData_Count$Cohort=="HC" & EukPresentData_Count$DisplayWeek=="24 Weeks")]/38


euk_time_combo <- unique(expand.grid(Ent_Blast_data$DisplayWeek, Ent_Blast_data_long$Organism))

colnames(euk_time_combo) <- c("DisplayWeek","Organism")

Euk_stats <- c() 

Euk_stats_pre <- rbind(c("Naïve", "Exp"),
                       c("Naïve", "HC"),
                       c("Exp","HC"))

for (r in 1:nrow(euk_time_combo)) {

  org_r <- as.character(euk_time_combo$Organism[r])
  dw_r <- as.character(euk_time_combo$DisplayWeek[r])
  
  Euk_stats_r <- cbind(Euk_stats_pre, rep(org_r,3),rep(dw_r,3),rep(NA,3),rep(NA,3))  
  
  Euk_stats_r <- as.data.frame(Euk_stats_r)
  
  colnames(Euk_stats_r) <- c("group1","group2","Organism","DisplayWeek","pval","padj")
  
  for (r2 in 1:nrow(Euk_stats_r)) {
  
    data_r_r2 <- Ent_Blast_data[which(Ent_Blast_data$Cohort %in% c(Euk_stats_r$group1[r2],Euk_stats_r$group2[r2]) & Ent_Blast_data$DisplayWeek==dw_r),c("Cohort",org_r)]
    colnames(data_r_r2) <- c("Cohort","Present")
    
    chi_r_r2 <- chisq_test(x=data_r_r2$Cohort,data_r_r2$Present)
    
    Euk_stats_r$pval[r2] <- chi_r_r2$p
  
  }
  
  Euk_stats_r$padj <- p.adjust(Euk_stats_r$pval, method="bonferroni")
  
  Euk_stats <- rbind(Euk_stats, Euk_stats_r)

}

EukPresentData_Count$Organism <- factor(EukPresentData_Count$Organism,levels=c("Blastocystis","Entamoeba","Komagataella" ,"Strigeidida"))

Euk_stats$Organism <- factor(Euk_stats$Organism,levels=c("Blastocystis","Entamoeba","Komagataella","Strigeidida"))

Euk_stats$group1 <- factor(Euk_stats$group1,levels=c("Naïve", "Exp", "HC"))

Euk_stats$group2 <- factor(Euk_stats$group2,levels=c("Naïve", "Exp", "HC"))

Euk_stats$Cohort <- Euk_stats$group1

Euk_stats$p.signif <- add_asterisks(Euk_stats$padj)

Euk_stats$xmin <- as.numeric(Euk_stats$Organism) + (as.numeric(Euk_stats$group1)-2)*.25
Euk_stats$xmax <- as.numeric(Euk_stats$Organism) + (as.numeric(Euk_stats$group2)-2)*.25


             

Euk_barplot <- ggplot(data=EukPresentData_Count, aes(x=Organism, y=Percent, fill=Cohort))+
                geom_bar(stat="identity", position="dodge")+
                facet_nested(rows=vars(DisplayWeek))+
                stat_pvalue_manual(
                  Euk_stats,
                  label = "p.signif",
                  hide.ns = TRUE,
                  y.position = .8,
                  step.increase = .1,
                  step.group.by = "Organism",
                )

BlastPresentData_Count <- EukPresentData_Count[which(as.character(EukPresentData_Count$Organism)=="Blastocystis"),]


test <- Ent_Blast_data[which(Ent_Blast_data$Blastocystis!=""),] 
Ent_Blast_data_simp <- Ent_Blast_data[which(Ent_Blast_data$Blastocystis!=""),c("PID","Cohort","Treatment","HIV_Status","DisplayWeek","Blastocystis","HDL","LDL","Glucose","Trig")]
Blast_data_wide <- pivot_wider(Ent_Blast_data_simp, 
                               id_cols = c("PID","Cohort"),
                               names_from="DisplayWeek",
                               values_from="Blastocystis")

n_Blast_data_wide <- Blast_data_wide[which(Blast_data_wide$Cohort=="Naïve"),]

n_blast_table <- table(n_Blast_data_wide$`0 Weeks`, n_Blast_data_wide$`24 Weeks`)

e_Blast_data_wide <- Blast_data_wide[which(Blast_data_wide$Cohort=="Exp"),]

e_blast_table <- table(e_Blast_data_wide$`0 Weeks`, e_Blast_data_wide$`24 Weeks`)
  
h_Blast_data_wide <- Blast_data_wide[which(Blast_data_wide$Cohort=="HC"),]
  
h_blast_table <- table(h_Blast_data_wide$`0 Weeks`, h_Blast_data_wide$`24 Weeks`)



n_mcnemar_test <- as.data.frame(mcnemar_test(n_blast_table))
n_mcnemar_test$Cohort<-"Naïve"
n_mcnemar_test$group1<-"0 Weeks"
n_mcnemar_test$group2<-"24 Weeks"


e_mcnemar_test <- as.data.frame(mcnemar_test(e_blast_table))
e_mcnemar_test$Cohort<-"Exp"
e_mcnemar_test$group1<-"0 Weeks"
e_mcnemar_test$group2<-"24 Weeks"

h_mcnemar_test <- as.data.frame(mcnemar_test(h_blast_table))
h_mcnemar_test$Cohort<-"HC"
h_mcnemar_test$group1<-"0 Weeks"
h_mcnemar_test$group2<-"24 Weeks"



Blast_tm_stats <- rbind(n_mcnemar_test, e_mcnemar_test, h_mcnemar_test)

Blast_tm_stats$Cohort <- factor(Blast_tm_stats$Cohort,levels=c("Naïve", "Exp", "HC"))

Blast_co_stats <- Euk_stats[which(Euk_stats$Organism=="Blastocystis"),]

Blast_co_stats$DisplayWeek <- factor(Blast_co_stats$DisplayWeek, levels=c("0 Weeks","24 Weeks"))

Blast_co_stats$xmin <- as.numeric(Blast_co_stats$DisplayWeek) + (as.numeric(Blast_co_stats$group1)-2)*.3
Blast_co_stats$xmax <- as.numeric(Blast_co_stats$DisplayWeek) + (as.numeric(Blast_co_stats$group2)-2)*.3

Blast_tm_stats$xmin <- 1 + (as.numeric(Blast_tm_stats$Cohort)-2)*.3
Blast_tm_stats$xmax <- 2 + (as.numeric(Blast_tm_stats$Cohort)-2)*.3


# First I use them same color scheme as described in Zim1 
cohortColors <- palette(brewer.pal(n = 5, name = "BrBG"))
cohortColors <- palette(brewer.pal(n = 5, name = "BrBG")) #Repeated to solidify? 
visitColors <- palette(brewer.pal(n = 5, name = "PiYG")) 
visitColors <- palette(brewer.pal(n = 5, name = "PiYG")) #repeat to solidify

Blast_presence_plot <- ggplot(data=BlastPresentData_Count, aes(x=DisplayWeek, y=Percent, fill=Cohort))+
                geom_bar(stat="identity", position="dodge", color="black")+
                scale_fill_manual(values=c(cohortColors[1],
                                           "gray",
                                           cohortColors[5]))+
                labs(y="Percent of Samples Containing Blastocystsis", tag="A")+
                theme_classic() +
                theme(legend.position = "bottom",
                      legend.title = element_text(size = 20),
                      legend.text = element_text(size = 20),
                      axis.text.x = element_text(size = 20),
                      axis.title.y = element_text(size = 20),
                      axis.title.x = element_blank(),
                      axis.text.y = element_text(size = 15),
                      strip.text.x = element_text(size=15),
                      strip.text.y = element_text(size=15),
                      plot.tag = element_text(size=25),
                      )+
                stat_pvalue_manual(
                  Blast_tm_stats,
                  label = "p.signif",
                  hide.ns = TRUE,
                  y.position = 1,
                  step.increase = .1
                )+
                stat_pvalue_manual(
                  Blast_co_stats,
                  label = "p.signif",
                  hide.ns = TRUE,
                  y.position = 0.75,
                  step.increase = .1
                )



```

### Linear Mixed Effects Models  - All Participants

```{r}

met_vars <- c("HDL","LDL","Trig","Glucose")
blast_vars <- "Blastocystis"

#We generate an linear mixed effects model data set to inlcude results
blast_lm_data <- unique(expand.grid(blast_vars, met_vars))

colnames(blast_lm_data) <- c("Predictor","Outcome")
#We generate fields for the statistics
blast_lm_data["pval"] <- NA
blast_lm_data["padj"] <- NA
blast_lm_data["coef"] <- NA
blast_lm_data["sig"] <- NA
blast_lm_data["percent_variance_exp"] <- NA

Ent_Blast_data$Blastocystis <- factor(Ent_Blast_data$Blastocystis, levels=c("No","Yes"), ordered = TRUE)
for (response in met_vars) { 
  #We pull the confounders for that variable
  conf_r <- get(paste0("conf_vars_",response))
  
  #We add them to the blast_vars
  blast_vars_full <- c(blast_vars,conf_r)
  
  #We remove missing variables
  data_r <- na.omit(Ent_Blast_data[,c(response,blast_vars_full,"PID")])
  
  # create full model 
  ff <- as.formula(paste0(response,"~",paste0(blast_vars_full, collapse="+"),"+(1|PID)")) 
  fm <- lmer(ff,
             data = data_r)
  
  predict_fm <- predict(fm)
  sum_fm <- summary(fm)
  
  #Below we calculate the R -suared value
  #First total sum of squares
  tss <- sum((unlist(data_r[,c(response)])-mean(unlist(data_r[,c(response)])))^2)
  #Second residual sum of squares
  rss <- sum((unlist(data_r[,c(response)])-predict_fm)^2)
  #R-squared
  r_squared <- 1-(rss/tss)
  #Percent variance explained
  pve <- r_squared*100
  
  blast_lm_data$percent_variance_exp[which(blast_lm_data$Outcome==response)] <- pve
  class(pve) <- "numeric"
  
  #We pull out the coefficinets
  coef_fm <- as.data.frame(sum_fm$coefficients)
  coef_fm$padj <- p.adjust(coef_fm$`Pr(>|t|)`, method="fdr")
  
  #Now we go through each predictor and pull out the coefficient in p-value 
  for (pred in blast_vars) {
    #We pull the row in the coefficient data set (use substr because sometime strings are added)
    coef_row <- which(substr(rownames(coef_fm),1,nchar(pred))==pred)
    #We not pull the row from the lm data set to put it in
    stat_row <- which(as.character(blast_lm_data$Predictor)==pred & as.character(blast_lm_data$Outcome)==response)
    #We add an if statemenbt to make sure we only add values when its ad 
    
    if (length(coef_row)==1 & length(stat_row)==1) {
      #We pull out the p-valueand adjusted p-value
      blast_lm_data$pval[stat_row] <- coef_fm$`Pr(>|t|)`[coef_row]
      blast_lm_data$padj[stat_row] <- coef_fm$padj[coef_row]
      #We pull out the coefficient
      blast_lm_data$coef[stat_row] <- coef_fm$Estimate[coef_row]
      
      #Based on the direction we indicate if it is positive of negative
      if (blast_lm_data$coef[stat_row]>0) {
        blast_lm_data$sig[stat_row] <- "Positive"
      } 
      if (blast_lm_data$coef[stat_row]<0) {
        blast_lm_data$sig[stat_row] <- "Negative"
      } 
      if (blast_lm_data$padj[stat_row]<0.05) {
        blast_lm_data$sig[stat_row] <- paste0(blast_lm_data$sig[stat_row], " (Significant)")
      }
      
      
      
    }
  }
  
  
}


#We add -C to the LDL and HDL for visualization
blast_lm_data$DisplayOutcome <- gsub("DL","DL-C",blast_lm_data$Outcome)
#We replace the trig abbreviation with the full Triglycerides name 
blast_lm_data$DisplayOutcome <- gsub("Trig","Triglycerides",blast_lm_data$DisplayOutcome)
#Now we add the percent variance explained
blast_lm_data$DisplayOutcome <- paste0(blast_lm_data$DisplayOutcome, " (",round(blast_lm_data$percent_variance_exp,1),"% exp)")






blast_plot <- ggplot(blast_lm_data, aes(x = Predictor, y = DisplayOutcome, color=sig)) +
  scale_color_manual(values=c("lightblue","pink","red"))+
  geom_point(size=5) +
  theme_minimal() +
  labs(x = "Plasma Measure", y = " ", fill=" ", tag="B", title="All Participants") +
  theme_classic() +
  guides(color = guide_legend(ncol = 1))+
  theme(legend.position = "right",
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        axis.text.y = element_text(size=12),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(size = 12, angle=45, hjust=1),
        strip.text.x = element_text(size=15),
        strip.text.y = element_text(size=15),
        title = element_text(size=20),
        plot.tag = element_text(size=25))


```

### Linear Mixed Effects Models  - Naive Participants
```{r}

Ent_Blast_data_naive <- Ent_Blast_data[which(Ent_Blast_data$Cohort=="Naïve"),]
#We generate an linear mixed effects model data set to inlcude results
blast_lm_data_naive <- unique(expand.grid(blast_vars, met_vars))

colnames(blast_lm_data_naive) <- c("Predictor","Outcome")
#We generate fields for the statistics
blast_lm_data_naive["pval"] <- NA
blast_lm_data_naive["padj"] <- NA
blast_lm_data_naive["coef"] <- NA
blast_lm_data_naive["sig"] <- NA
blast_lm_data_naive["percent_variance_exp"] <- NA

for (response in met_vars) { 
  #We pull the confounders for that variable
  conf_r <- get(paste0("conf_vars_",response))
  
  #We add them to the blast_vars
  #Because they are all HIV positive (HIV should be removed if in there)
  
  blast_vars_full <- setdiff(c(blast_vars,conf_r),"HIV_Status")
  
  #We remove missing variables
  data_r <- na.omit(Ent_Blast_data_naive[,c(response,blast_vars_full,"PID")])
  
  # create full model 
  ff <- as.formula(paste0(response,"~",paste0(blast_vars_full, collapse="+"),"+(1|PID)")) 
  fm <- lmer(ff,
             data = data_r)
  
  predict_fm <- predict(fm)
  sum_fm <- summary(fm)
  
  #Below we calculate the R -suared value
  #First total sum of squares
  tss <- sum((unlist(data_r[,c(response)])-mean(unlist(data_r[,c(response)])))^2)
  #Second residual sum of squares
  rss <- sum((unlist(data_r[,c(response)])-predict_fm)^2)
  #R-squared
  r_squared <- 1-(rss/tss)
  #Percent variance explained
  pve <- r_squared*100
  
  blast_lm_data_naive$percent_variance_exp[which(blast_lm_data_naive$Outcome==response)] <- pve
  class(pve) <- "numeric"
  
  #We pull out the coefficinets
  coef_fm <- as.data.frame(sum_fm$coefficients)
  coef_fm$padj <- p.adjust(coef_fm$`Pr(>|t|)`, method="fdr")
  
  #Now we go through each predictor and pull out the coefficient in p-value 
  for (pred in blast_vars) {
    #We pull the row in the coefficient data set (use substr because sometime strings are added)
    coef_row <- which(substr(rownames(coef_fm),1,nchar(pred))==pred)
    #We not pull the row from the lm data set to put it in
    stat_row <- which(as.character(blast_lm_data_naive$Predictor)==pred & as.character(blast_lm_data_naive$Outcome)==response)
    #We add an if statemenbt to make sure we only add values when its ad 
    
    if (length(coef_row)==1 & length(stat_row)==1) {
      #We pull out the p-valueand adjusted p-value
      blast_lm_data_naive$pval[stat_row] <- coef_fm$`Pr(>|t|)`[coef_row]
      blast_lm_data_naive$padj[stat_row] <- coef_fm$padj[coef_row]
      #We pull out the coefficient
      blast_lm_data_naive$coef[stat_row] <- coef_fm$Estimate[coef_row]
      
      #Based on the direction we indicate if it is positive of negative
      if (blast_lm_data_naive$coef[stat_row]>0) {
        blast_lm_data_naive$sig[stat_row] <- "Positive"
      } 
      if (blast_lm_data_naive$coef[stat_row]<0) {
        blast_lm_data_naive$sig[stat_row] <- "Negative"
      } 
      if (blast_lm_data_naive$padj[stat_row]<0.05) {
        blast_lm_data_naive$sig[stat_row] <- paste0(blast_lm_data_naive$sig[stat_row], " (Significant)")
      }
      
      
      
    }
  }
  
  
}


#We add -C to the LDL and HDL for visualization
blast_lm_data_naive$DisplayOutcome <- gsub("DL","DL-C",blast_lm_data_naive$Outcome)
#We replace the trig abbreviation with the full Triglycerides name 
blast_lm_data_naive$DisplayOutcome <- gsub("Trig","Triglycerides",blast_lm_data_naive$DisplayOutcome)
#Now we add the percent variance explained
blast_lm_data_naive$DisplayOutcome <- paste0(blast_lm_data_naive$DisplayOutcome, " (",round(blast_lm_data_naive$percent_variance_exp,1),"% exp)")


blast_plot_naive <- ggplot(blast_lm_data_naive, aes(x = Predictor, y = DisplayOutcome, color=sig)) +
  scale_color_manual(values=c("lightblue","blue","pink","red"))+
  geom_point(size=5) +
  theme_minimal() +
  labs(x = "Plasma Measure", y = " ", fill=" ", tag="C", title="Naïve Participants") +
  theme_classic() +
  guides(color = guide_legend(ncol = 1))+
  theme(legend.position = "right",
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        axis.text.y = element_text(size=12),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(size = 12, angle=45, hjust=1),
        strip.text.x = element_text(size=15),
        strip.text.y = element_text(size=15),
        title = element_text(size=20),
        plot.tag = element_text(size=25))


```

###Delta Boxplot
```{r}

Ent_Blast_data_lmer <- Ent_Blast_data[which(!is.na(Ent_Blast_data$Blastocystis)),]
Blast_delta_base <- unique(Ent_Blast_data_lmer[,c("PID","Cohort", "Treatment", "HIV_Status")])

Blast_delta <- Blast_delta_base

delta_vars <- c("Blastocystis","HDL","LDL","Trig","Glucose")

for (v in delta_vars) {
 Blast_delta$V <- NA
 for (r in 1:nrow(Blast_delta)) {
   pid_r <- Blast_delta$PID[r]
   row_0 <- which(Ent_Blast_data_lmer$SampleID==paste0(pid_r,".2"))
   row_24 <- which(Ent_Blast_data_lmer$SampleID==paste0(pid_r,".3"))
   if (length(row_0)==1 & length(row_24)==1) {
     if (v=="Blastocystis") {
       if (Ent_Blast_data_lmer$Blastocystis[row_0]=="Yes" & Ent_Blast_data_lmer$Blastocystis[row_24]=="No") {
         delta_r <- "Lost"
       } 
       if (Ent_Blast_data_lmer$Blastocystis[row_0]=="Yes" & Ent_Blast_data_lmer$Blastocystis[row_24]=="Yes") {
         delta_r <- "Present"
         #delta_r <- "Other"
       }
      if (Ent_Blast_data_lmer$Blastocystis[row_0]=="No" & Ent_Blast_data_lmer$Blastocystis[row_24]=="Yes") {
         delta_r <- "Acquired"
        # delta_r <- "Other"
      }
      if (Ent_Blast_data_lmer$Blastocystis[row_0]=="No" & Ent_Blast_data_lmer$Blastocystis[row_24]=="No") {
          delta_r <- "Absent"

       }
     } else {
       delta_r <- as.numeric(Ent_Blast_data_lmer[row_24,v]) - as.numeric(Ent_Blast_data_lmer[row_0,v])
     }
     Blast_delta$V[r] <- delta_r
   }
 }
 colnames(Blast_delta)[ncol(Blast_delta)] <- paste0("delta_",v)
}


Blast_delta_filt <- Blast_delta[which(!is.na(Blast_delta$delta_Blastocystis)),]

Blast_delta_long <- pivot_longer(data=Blast_delta_filt,
                                 cols=c("delta_HDL","delta_LDL","delta_Trig", "delta_Glucose"),
                                 names_to = "Marker",
                                 values_to = "delta")






Blast_delta_stat <- gp_stat_sum(data=Blast_delta_long,
                                 sub_vars=c("Marker"),
                                 dep_vars=c("delta"),
                                 gp_var="delta_Blastocystis",
                                 normality= FALSE,  
                                 adj="bonferroni"    
                                 )   

Blast_delta_long_naive <- Blast_delta_long[which(Blast_delta_long$Cohort=="Naïve"),]

Blast_delta_stat_naive <- gp_stat_sum(data=Blast_delta_long_naive,
                                 sub_vars=c("Marker"),
                                 dep_vars=c("delta"),
                                 gp_var="delta_Blastocystis",
                                 normality= FALSE,  
                                 adj="bonferroni"    
                                 )   

Blast_delta_stat_naive$Cohort <- "Naïve"

Blast_delta_long_hc <- Blast_delta_long[which(Blast_delta_long$Cohort=="HC"),]

Blast_delta_stat_hc <- gp_stat_sum(data=Blast_delta_long_hc,
                                 sub_vars=c("Marker"),
                                 dep_vars=c("delta"),
                                 gp_var="delta_Blastocystis",
                                 normality= FALSE,  
                                 adj="bonferroni"    
                                 )   

Blast_delta_stat_hc$Cohort="HC" 

Blast_delta_long_exp <- Blast_delta_long[which(Blast_delta_long$Cohort=="Exp"),]

Blast_delta_stat_exp <- gp_stat_sum(data=Blast_delta_long_exp,
                                 sub_vars=c("Marker"),
                                 dep_vars=c("delta"),
                                 gp_var="delta_Blastocystis",
                                 normality= FALSE,  
                                 adj="bonferroni"    
                                 )   

Blast_delta_stat_exp$Cohort <- "Exp"

Blast_delta_stats_all <- rbind(Blast_delta_stat_exp, Blast_delta_stat_hc, Blast_delta_stat_naive)

Blast_delta_stats_all$Cohort <- factor(Blast_delta_stats_all$Cohort, levels=c("Naïve","Exp","HC"), ordered = TRUE)

Blast_delta_stats_all$delta_Blastocystis <- Blast_delta_stats_all$group1

Blast_delta_stats_all$xmin = 2.75
Blast_delta_stats_all$xmax = 3

Blast_delta_stats_all$DisplayMarker <- gsub("delta_","",Blast_delta_stats_all$Marker)
Blast_delta_stats_all$DisplayMarker <- gsub("DL","DL-C",Blast_delta_stats_all$DisplayMarker)
Blast_delta_stats_all$DisplayMarker <- gsub("Trig","Triglycerides",Blast_delta_stats_all$DisplayMarker)

Blast_delta_long$DisplayMarker <- gsub("delta_","",Blast_delta_long$Marker)
Blast_delta_long$DisplayMarker <- gsub("DL","DL-C",Blast_delta_long$DisplayMarker)
Blast_delta_long$DisplayMarker <- gsub("Trig","Triglycerides",Blast_delta_long$DisplayMarker)

delta_blast_box <- ggplot(data = Blast_delta_long,
                      mapping = aes(x = DisplayMarker, y = delta, fill=delta_Blastocystis)) +
                      geom_boxplot(aes(x=DisplayMarker, y = delta), 
                                   outliers = FALSE) +
                   #   geom_jitter(aes(x = DisplayMarker, y = delta, fill=delta_Blastocystis)) +
                      labs(y="Change in Measure (mg/dL)", tag="E", fill="Blastocystis")+
                      theme_classic() +
                      theme(legend.position = "right",
                            legend.title = element_text(size = 20),
                            legend.text = element_text(size = 20),
                            axis.text.x = element_text(size = 15, hjust=1, angle=45),
                            axis.title.y = element_text(size = 20),
                            axis.title.x = element_blank(),
                            axis.text.y = element_text(size = 15),
                            strip.text.x = element_text(size=15),
                            strip.text.y = element_text(size=10),
                            plot.tag = element_text(size=25),
                            ggh4x.facet.nestline = element_blank(),
                            strip.background.x = element_rect(color="black", fill="gray", linetype="solid"),
                            strip.background.y = element_rect(color="black", fill="gray", linetype="solid")) +
                      facet_nested(cols=vars(Cohort), 
                                  scales="free_y") 

Blast_delta_long_naive <- Blast_delta_long[which(Blast_delta_long$Cohort=="Naïve"),]
Blast_delta_stats_all_naive <- Blast_delta_stats_all[which(Blast_delta_stats_all$Cohort=="Naïve"),]

delta_blast_box_naive <- ggplot(data = Blast_delta_long_naive,
                      mapping = aes(x = DisplayMarker, y = delta, fill=delta_Blastocystis)) +
                      geom_boxplot(aes(x=DisplayMarker, y = delta), 
                                   outliers = FALSE) +
                      geom_jitter(aes(x = DisplayMarker, y = delta, fill=delta_Blastocystis)) +
                      labs(y="Change in Measure (mg/dL)", tag="F", fill="Blastocystis", title="Naïve Participants")+
                      theme_classic() +
                      theme(legend.position = "right",
                            legend.title = element_text(size = 25),
                            legend.text = element_text(size = 20),
                            title = element_text(size=20),
                            axis.text.x = element_text(size = 15, hjust=1, angle=45),
                            axis.title.y = element_text(size = 20),
                            axis.title.x = element_blank(),
                            axis.text.y = element_text(size = 15),
                            strip.text.x = element_text(size=15),
                            strip.text.y = element_text(size=10),
                            plot.tag = element_text(size=25),
                            ggh4x.facet.nestline = element_blank(),
                            strip.background.x = element_rect(color="black", fill="gray", linetype="solid"),
                            strip.background.y = element_rect(color="black", fill="gray", linetype="solid")) 

                      

delta_blast_box_all <- ggplot(data = Blast_delta_long,
                      mapping = aes(x = Marker, y = delta, fill=delta_Blastocystis)) +
                      geom_boxplot(aes(x=Marker, y = delta), 
                                   outliers = FALSE) +
                      geom_jitter(aes(x = Marker, y = delta)) +
                      labs(y="Change in Measure (mg/dL)", tag="C")+
                      theme_classic() +
                      theme(legend.position = "right",
                            legend.title = element_text(size = 20),
                            legend.text = element_text(size = 20),
                            axis.text.x = element_text(size = 15, hjust=1, angle=45),
                            axis.title.y = element_text(size = 20),
                            axis.title.x = element_blank(),
                            axis.text.y = element_text(size = 15),
                            strip.text.x = element_text(size=15),
                            strip.text.y = element_text(size=10),
                            plot.tag = element_text(size=25),
                            ggh4x.facet.nestline = element_blank(),
                            strip.background.x = element_rect(color="black", fill="gray", linetype="solid"),
                            strip.background.y = element_rect(color="black", fill="gray", linetype="solid")) 





```

###Blastocystis DAG
```{r}


# Nodes are created in a data frame
nodes_Blast <- data.frame(
  label = c("ART and\nCotrimoxazole\nTreatment", "Decreased\nBlastocystis", "Increased\nTriglycerides"),
  x = c(1, 2, 3),
  y = c(1, 2, 1),
  category=c("Predictor","Mediator","Outcome")
)

# Original arrow coordinates
arrows_Blast <- data.frame(
  x1 = c(1, 1, 2),
  y1 = c(1, 1, 2),
  x2 = c(2, 3, 3),
  y2 = c(2, 1, 1)
)

# Calculate new coordinates for the arrows

arrows_Blast_adjusted <- arrows_Blast
for (i in 1:nrow(arrows_Blast)) {
  x_diff <- arrows_Blast$x2[i] - arrows_Blast$x1[i]
  y_diff <- arrows_Blast$y2[i] - arrows_Blast$y1[i]
  length_orig <- sqrt(x_diff^2 + y_diff^2)
  if (length_orig > 0) {
    arrows_Blast_adjusted$x2[i] <- arrows_Blast$x1[i] + x_diff * ((length_orig - (length_orig/2)) / length_orig)
    arrows_Blast_adjusted$y2[i] <- arrows_Blast$y1[i] + y_diff * ((length_orig - (length_orig/2)) / length_orig)
  }
}

# Arrow Labels are created in a data frame
arrow_labels_Blast <- data.frame(
  x = c(1.3, 2, 2.7),
  y = c(1.6, 1.1, 1.6),
  label = c("-3.4**","10.6(5.9)","-15.5*")
)

# The DAG is made in GGplot with adjusted arrows
Blast_med_dag <- ggplot() +
  geom_segment(data = arrows_Blast_adjusted,  # Use the adjusted coordinates
               aes(x = x1, y = y1, xend = x2, yend = y2),
               arrow = arrow(length = unit(0.3, "cm"), type = "closed")) +
  geom_segment(data = arrows_Blast,  # Use the adjusted coordinates
               aes(x = x1, y = y1, xend = x2, yend = y2),
               arrow = arrow(length = unit(0.3, "cm"), type = "closed")) +
  geom_text(data = arrow_labels_Blast, aes(x = x, y = y, label = label), size = 4) +
  geom_label(data = nodes_Blast, 
             aes(x = x, y = y, label = label, fill = category), 
             size = 5, 
             key_glyph = "rect") +
  scale_fill_manual(values = c("orange", "red", "lightblue")) +
  scale_x_continuous(limits = c(0.5, 3.5)) +
  scale_y_continuous(limits = c(0.5, 2.5)) +
  theme_void() +
  labs(title = "Mediation Results", 
       tag = "D", 
       fill = " ") +
  theme(legend.position = "right",
        legend.title = element_text(size = 20),
        legend.text = element_text(size = 20),
        plot.tag = element_text(size = 25),
        title = element_text(size = 20))


```


###Plot Arrangement
```{r}

fig_BC <- ggarrange(blast_plot, blast_plot_naive, nrow=1)

fig_BCD <- ggarrange(fig_BC,Blast_med_dag, nrow=2)

fig_all <- ggarrange(Blast_presence_plot,fig_BCD, nrow=1, widths=c(1,2))
```

