---
title: "Baseline Variable Selection"
author: "J. O'Connor"
date: "2025-09-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Baseline Variable Selection

This markdown completed the baseline varbiable selection to identify potential confounders contributing to metabolic outcome

### Function Sourcing

First the functions are all sourced
```{r Functions, message=FALSE, warning=FALSE}
#To access the functions directory why must direct to the parent directory
setwd("../..")

#A list of the function files is pulled
function_files <- list.files(paste0(getwd(),"/Functions"), pattern = "\\.R$", full.names = TRUE)
#Those functions are then sourced
lapply(function_files, source)
```


### Data Loading

We load all the data of interest

```{r data_load, echo=FALSE}
#All Metadata objects have been stored in one location
metadata_objects <- readRDS("~/Zimbabwe-Cardiometabolic-Analysis/Data/R_Data/metadata_objects.RDS")

#Metdata objects
  #base: all processed metadata
  base <- metadata_objects$base

#For the analysis we only work with the basleine data
  
base_t1 <- base[which(base$Week=="0"),]


  

```


###Variable Definition and filtering

Outcomes and confounders are manually selected

```{r}

#Metabolic outcome variables are defined

#HDL: plasma high-density lipoprotein cholesterol (HDL-C)
#LDL: plasma low-density lipoprotein cholesterol (HDL-C)
#Trig: plasma triglycerides
#Glucose: Plasma glucose

met_vars <- c("HDL",
              "LDL",
              "Trig",
              "Glucose")

#All counfounders are inputted

#Visit_Age: Age in years at baseline
#Location: Urban or Rural clinic
#BMI: Body Mass Index
#Gender: Sex
#HIV_Status: HIV infection status
#Treatment: Indication if the subject was on treatment at baseline (distinguiding Exp and Naive groups)
#Dietary Information:
  #Total_Dairy
  #Total_Fruit
  #Total_Veg
  #Total_Protein
  #Total_Grain  
  #Total_Fast_Food
  #Total_Drinks
  #Total_Snacks


confs <- c("Visit_Age",
           "Location",
           "BMI",
           "Gender",
           "HIV_Status",
           "Treatment",
           "Total_Dairy",
           "Total_Fruit",
           "Total_Veg" ,
           "Total_Protein",
           "Total_Grain",
           "Total_Fast_Food",
           "Total_Drinks",
           "Total_Snacks"
           )


# 1. Identify which variables are categorical vs numerical
cat_vars <- confs[sapply(base_t1[confs], is.factor) | sapply(base_t1[confs], is.character)]
num_vars <- confs[sapply(base_t1[confs], is.numeric)]

# Initialize a results list
bivariate_results <- list()

# --- A. Spearman Correlation with P-Values (Numerical vs Numerical) ---
if(length(num_vars) > 1) {
  # Create all unique pairs of numerical variables
  num_combos <- expand.grid(Var1 = num_vars, Var2 = num_vars) %>%
    filter(as.character(Var1) < as.character(Var2)) # Avoid self-correlation and duplicates
  
  bivariate_results$num_num <- apply(num_combos, 1, function(row) {
    # cor.test provides the p-value
    test <- cor.test(base_t1[[row[1]]], base_t1[[row[2]]], 
                     method = "spearman", exact = FALSE)
    
    data.frame(
      Predictor_A = row[1], 
      Predictor_B = row[2], 
      Rho = test$estimate, 
      p_val = test$p.value
    )
  }) %>% bind_rows() %>% arrange(p_val)
}

# --- B. Chi-Square Test (Categorical vs Categorical) ---
cat_combos <- expand.grid(Var1 = cat_vars, Var2 = cat_vars) %>% 
  filter(as.character(Var1) < as.character(Var2))

bivariate_results$cat_cat <- apply(cat_combos, 1, function(row) {
  # Clean out missing data for the table
  tab <- table(base_t1[[row[1]]], base_t1[[row[2]]])
  
  # Only run if table has dimensions
  if(nrow(tab) > 1 & ncol(tab) > 1) {
    test <- chisq.test(tab)
    return(data.frame(Var1 = row[1], Var2 = row[2], p_val = test$p.value))
  }
  return(NULL)
}) %>% bind_rows() %>% arrange(p_val)

# --- C. Wilcoxon Rank Sum (Numerical vs Categorical) ---
mixed_combos <- expand.grid(Num = num_vars, Cat = cat_vars)

bivariate_results$num_cat <- apply(mixed_combos, 1, function(row) {
  # Ensure the category has exactly 2 levels for Wilcoxon
  lvls <- length(unique(na.omit(base_t1[[row[2]]])))
  
  if(lvls == 2) {
    test <- wilcox.test(as.formula(paste(row[1], "~", row[2])), data = base_t1)
    return(data.frame(Numeric = row[1], Category = row[2], p_val = test$p.value))
  } else if (lvls > 2) {
    # If more than 2 levels (e.g., Location with 3+ types), use Kruskal-Wallis
    test <- kruskal.test(as.formula(paste(row[1], "~", row[2])), data = base_t1)
    return(data.frame(Numeric = row[1], Category = row[2], p_val = test$p.value))
  }
  return(NULL)
}) %>% bind_rows() %>% arrange(p_val)


# --- 1. Numerical-Numerical (High Rho & Significant) ---
high_cor_num <- bivariate_results$num_num %>%
  filter(p_val < 0.05) %>%
  mutate(Type = "Num-Num (Spearman)")

# --- 2. Categorical-Categorical (Highly Associated) ---
high_cor_cat <- bivariate_results$cat_cat %>%
  filter(p_val < 0.01) %>% # Stricter p-value for Chi-square
  mutate(Type = "Cat-Cat (Chi-Square)")

# --- 3. Mixed Numerical-Categorical (Significant Differences) ---
high_cor_mixed <- bivariate_results$num_cat %>%
  filter(p_val < 0.05) %>%
  mutate(Type = "Num-Cat (Wilcox/Kruskal)")

# Display a consolidated 'Warning' table
warning_table <- bind_rows(
  high_cor_num %>% select(Var1 = Predictor_A, Var2 = Predictor_B, p_val, Type),
  high_cor_cat %>% select(Var1, Var2, p_val, Type),
  high_cor_mixed %>% select(Var1 = Numeric, Var2 = Category, p_val, Type)
)

print(warning_table)
#We do see correlations

# Quick check for Multicollinearity

# 1. Log Transformation & Scaling
base_t1_model <- base_t1 %>%
  mutate(across(where(is.ordered), as.factor)) # Convert ordered to regular factors

# Apply log transform to outcomes that failed normality
for(met in met_vars) {
  base_t1_model[[paste0(met, "_log")]] <- log10(base_t1_model[[met]] + 1)
}
met_vars_log <- paste0(met_vars, "_log")

# We use a dummy model with all confounders to see who is 'fighting'
hdl_vif_model <- lm(as.formula(paste("HDL_log", "~", paste(confs, collapse="+"))), data = base_t1_model)
car::vif(hdl_vif_model)

# We use a dummy model with all confounders to see who is 'fighting'
ldl_vif_model <- lm(as.formula(paste("LDL_log", "~", paste(confs, collapse="+"))), data = base_t1_model)
car::vif(ldl_vif_model)

# We use a dummy model with all confounders to see who is 'fighting'
glucose_vif_model <- lm(as.formula(paste("Glucose_log", "~", paste(confs, collapse="+"))), data = base_t1_model)
car::vif(glucose_vif_model)

# We use a dummy model with all confounders to see who is 'fighting'
trig_vif_model <- lm(as.formula(paste("Trig_log", "~", paste(confs, collapse="+"))), data = base_t1_model)
car::vif(trig_vif_model)

#Variance Inflation Factors are reasonable so do not need to move forward with limiting these 

```

#Data Split

This chunk prepares the data for all subsequent modeling sections.
```{r}

# Scale numeric predictors
num_confs <- confs[sapply(base_t1_model[,confs], is.numeric)]
base_t1_model[, num_confs] <- scale(base_t1_model[, num_confs])

# 2. Global Test/Train Split (ensures all methods use the same data)
# We store these in a list keyed by the response name
data_splits <- list()

for (response in met_vars_log) {
  working_data <- na.omit(base_t1_model[, c(confs, response)])
  
  set.seed(123)
  train_idx <- createDataPartition(working_data[[response]], p = 0.7, list = FALSE)
  
  data_splits[[response]] <- list(
    train = working_data[train_idx, ],
    test  = working_data[-train_idx, ]
  )
}
```



####Stepwise Selection 

#####Backward

```{r}
# Initialize results for this section
performance_bw <- data.frame()
selection_data_bw <- data.frame()
plots_importance_bw <- list()
plots_pred_obs_bw <- list()

for (response in met_vars_log) {
  # Pull the specific split for this response
  train_df <- data_splits[[response]]$train
  test_df  <- data_splits[[response]]$test
  
  # 1. Fit the Model
  full_formula <- as.formula(paste0(response, " ~ ."))
  full_lm <- lm(full_formula, data = train_df)
  bw_model <- step(full_lm, direction = "backward", trace = 0)
  
  # 2. Predictions & Performance
  bw_preds <- predict(bw_model, newdata = test_df)
  performance_bw <- rbind(performance_bw, 
                          get_metrics(test_df[[response]], bw_preds, "Backward", response))
  
  # 3. Importance Data (Coefficients)
  # Clean names to map 'GenderMale' -> 'Gender'
  coef_raw <- as.data.frame(summary(bw_model)$coefficients)
  colnames(coef_raw) <- c("Estimate", "StdErr", "t_val", "p_val")
  
  coef_df <- coef_raw %>%
    filter(rownames(.) != "(Intercept)") %>%
    mutate(
      Predictor = clean_pred_names(rownames(.), confs),
      Outcome = response,
      Direction = ifelse(Estimate > 0, "Positive", "Negative"),
      Sig = ifelse(p_val < 0.05, "Significant", "Non-Sig")
    )
  selection_data_bw <- rbind(selection_data_bw, coef_df)
  
  # 4. Generate Plots
  # Importance Plot
  plots_importance_bw[[response]] <- ggplot(coef_df, aes(x = reorder(Predictor, Estimate), y = Estimate, fill = Direction)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme_minimal() +
    labs(title = paste("BW Importance:", response), x = "Confounder", y = "Standardized Coefficient")
  
  # Pred vs Obs Plot
  pred_obs_df <- data.frame(Observed = test_df[[response]], Predicted = bw_preds)
  plots_pred_obs_bw[[response]] <- ggplot(pred_obs_df, aes(x = Observed, y = Predicted)) +
    geom_point(color = "blue", alpha = 0.6) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    theme_classic() +
    labs(title = paste("BW Pred vs Obs:", response), subtitle = paste("R2 =", round(tail(performance_bw$R2, 1), 3)))
}

# Display Performance
print(performance_bw)

```

#####Forward
```{r}

# Initialize results for this section
performance_fw <- data.frame()
selection_data_fw <- data.frame()
plots_importance_fw <- list()
plots_pred_obs_fw <- list()

for (response in met_vars_log) {
  # Pull the specific split for this response
  train_df <- data_splits[[response]]$train
  test_df  <- data_splits[[response]]$test
  
  # 1. Fit the Model (Forward requires a null model and a scope)
  null_model <- lm(as.formula(paste0(response, " ~ 1")), data = train_df)
  full_formula <- as.formula(paste0(response, " ~ ."))
  
  # Perform Forward Selection
  fw_model <- step(null_model, 
                   scope = list(lower = null_model, upper = full_formula), 
                   direction = "forward", 
                   trace = 0)
  
  # 2. Predictions & Performance (Testing on holdout set)
  fw_preds <- predict(fw_model, newdata = test_df)
  performance_fw <- rbind(performance_fw, 
                          get_metrics(test_df[[response]], fw_preds, "Forward", response))
  
  # 3. Importance Data (Coefficients)
  # Clean names to map 'GenderMale' -> 'Gender'
  coef_raw <- as.data.frame(summary(fw_model)$coefficients)
  colnames(coef_raw) <- c("Estimate", "StdErr", "t_val", "p_val")
  
  coef_df <- coef_raw %>%
    filter(rownames(.) != "(Intercept)") %>%
    mutate(
      Predictor = clean_pred_names(rownames(.), confs),
      Outcome = response,
      Direction = ifelse(Estimate > 0, "Positive", "Negative"),
      Sig = ifelse(p_val < 0.05, "Significant", "Non-Sig")
    )
  selection_data_fw <- rbind(selection_data_fw, coef_df)
  
  # 4. Generate Plots
  # Importance Plot
  plots_importance_fw[[response]] <- ggplot(coef_df, aes(x = reorder(Predictor, Estimate), y = Estimate, fill = Direction)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme_minimal() +
    labs(title = paste("FW Importance:", response), x = "Confounder", y = "Standardized Coefficient")
  
  # Pred vs Obs Plot
  pred_obs_df <- data.frame(Observed = test_df[[response]], Predicted = fw_preds)
  plots_pred_obs_fw[[response]] <- ggplot(pred_obs_df, aes(x = Observed, y = Predicted)) +
    geom_point(color = "darkgreen", alpha = 0.6) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    theme_classic() +
    labs(title = paste("FW Pred vs Obs:", response), subtitle = paste("R2 =", round(tail(performance_fw$R2, 1), 3)))
}

# Display Performance
print(performance_fw)
```

#####Both

```{r}
# Initialize results for this section
performance_both <- data.frame()
selection_data_both <- data.frame()
plots_importance_both <- list()
plots_pred_obs_both <- list()

for (response in met_vars_log) {
  # Pull the specific split for this response
  train_df <- data_splits[[response]]$train
  test_df  <- data_splits[[response]]$test
  
  # 1. Fit the Model 
  # Stepwise 'both' usually starts from a full or null model; 
  # we'll start with full to allow for immediate subtraction/addition
  full_formula <- as.formula(paste0(response, " ~ ."))
  full_lm <- lm(full_formula, data = train_df)
  
  # Perform Stepwise Selection (Both directions)
  both_model <- step(full_lm, direction = "both", trace = 0)
  
  # 2. Predictions & Performance (Testing on holdout set)
  both_preds <- predict(both_model, newdata = test_df)
  performance_both <- rbind(performance_both, 
                           get_metrics(test_df[[response]], both_preds, "Both", response))
  
  # 3. Importance Data (Coefficients)
  # Clean names to map 'GenderMale' -> 'Gender'
  coef_raw <- as.data.frame(summary(both_model)$coefficients)
  colnames(coef_raw) <- c("Estimate", "StdErr", "t_val", "p_val")
  
  coef_df <- coef_raw %>%
    filter(rownames(.) != "(Intercept)") %>%
    mutate(
      Predictor = clean_pred_names(rownames(.), confs),
      Outcome = response,
      Direction = ifelse(Estimate > 0, "Positive", "Negative"),
      Sig = ifelse(p_val < 0.05, "Significant", "Non-Sig")
    )
  selection_data_both <- rbind(selection_data_both, coef_df)
  
  # 4. Generate Plots
  # Importance Plot
  plots_importance_both[[response]] <- ggplot(coef_df, aes(x = reorder(Predictor, Estimate), y = Estimate, fill = Direction)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme_minimal() +
    labs(title = paste("Stepwise (Both) Importance:", response), x = "Confounder", y = "Standardized Coefficient")
  
  # Pred vs Obs Plot
  pred_obs_df <- data.frame(Observed = test_df[[response]], Predicted = both_preds)
  plots_pred_obs_both[[response]] <- ggplot(pred_obs_df, aes(x = Observed, y = Predicted)) +
    geom_point(color = "purple", alpha = 0.6) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    theme_classic() +
    labs(title = paste("Both Pred vs Obs:", response), subtitle = paste("R2 =", round(tail(performance_both$R2, 1), 3)))
}

# Display Performance
print(performance_both)

```

###Lasso
```{r}
# Initialize results for this section
performance_lasso <- data.frame()
selection_data_lasso <- data.frame()
plots_importance_lasso <- list()
plots_pred_obs_lasso <- list()

# 1. Create master fold IDs outside the loop for 100% reproducibility
# This ensures cv.glmnet uses the same data splits every time you run the code
set.seed(527342)
# We use a large number (1000) or nrow(data) to ensure we have enough IDs for any split
fold_ids_master <- sample(rep(1:5, length.out = 2000)) 

for (response in met_vars_log) {
  # Pull the specific split
  train_df <- data_splits[[response]]$train
  test_df  <- data_splits[[response]]$test
  
  # Prepare matrices (Lasso requirement)
  # We use the formula to ensure dummy coding matches the other models
  x_train <- model.matrix(as.formula(paste(response, "~ .")), train_df)[,-1]
  y_train <- train_df[[response]]
  x_test  <- model.matrix(as.formula(paste(response, "~ .")), test_df)[,-1]
  y_test  <- test_df[[response]]
  
  # Subset master folds to match the current training rows
  current_fold_ids <- fold_ids_master[1:nrow(x_train)]
  
  # 2. Fit LASSO with Cross-Validation using fixed folds
  cv_lasso <- cv.glmnet(x_train, y_train, 
                        alpha = 1, 
                        foldid = current_fold_ids, 
                        standardize = TRUE)
  
  best_lam <- cv_lasso$lambda.min
  
  # 3. Extract selected variables
  lasso_coefs <- coef(cv_lasso, s = "lambda.min")
  selected_indices <- which(lasso_coefs != 0)[-1] # Exclude intercept
  selected_vars_raw <- rownames(lasso_coefs)[selected_indices]
  
  # 4. Refit and Performance Logic
  if(length(selected_vars_raw) > 0) {
    
    # Create formula for refit
    refit_formula <- as.formula(paste(response, "~", paste(selected_vars_raw, collapse = " + ")))
    
    # Explicitly build dataframes for the 'lm' refit to avoid 'object not found' errors
    refit_train_df <- as.data.frame(x_train[, selected_vars_raw, drop = FALSE])
    refit_train_df[[response]] <- y_train
    
    refit_test_df <- as.data.frame(x_test[, selected_vars_raw, drop = FALSE])
    refit_test_df[[response]] <- y_test
    
    # Refit Linear Model
    lasso_refit_model <- lm(refit_formula, data = refit_train_df)
    
    # 5. Predictions & Performance
    lasso_preds <- predict(lasso_refit_model, newdata = refit_test_df)
    performance_lasso <- rbind(performance_lasso, 
                               get_metrics(y_test, lasso_preds, "Lasso", response))
    
    # 6. Importance Data (Coefficients)
    coef_raw <- as.data.frame(summary(lasso_refit_model)$coefficients)
    colnames(coef_raw) <- c("Estimate", "StdErr", "t_val", "p_val")
    
    coef_df <- coef_raw %>%
      filter(rownames(.) != "(Intercept)") %>%
      mutate(
        Predictor = clean_pred_names(rownames(.), confs),
        Outcome = response,
        Direction = ifelse(Estimate > 0, "Positive", "Negative"),
        Sig = ifelse(p_val < 0.05, "Significant", "Non-Sig")
      )
    selection_data_lasso <- rbind(selection_data_lasso, coef_df)
    
    # 7. Generate Plots
    pred_obs_df <- data.frame(Observed = y_test, Predicted = as.vector(lasso_preds))
    plots_pred_obs_lasso[[response]] <- ggplot(pred_obs_df, aes(x = Observed, y = Predicted)) +
      geom_point(color = "darkorange", alpha = 0.6) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
      theme_classic() +
      labs(title = paste("Lasso Pred vs Obs:", response), 
           subtitle = paste("R2 =", round(tail(performance_lasso$R2, 1), 3)))
      
  } else {
  # Instead of just NA, calculate the error of the 'Mean' (Null Model)
  null_rmse <- sqrt(mean((y_test - mean(y_train))^2))
  null_mae <- mean(abs(y_test - mean(y_train)))
  
  performance_lasso <- rbind(performance_lasso, 
                             data.frame(Outcome = response, Method = "Lasso", 
                                        RMSE = null_rmse, MAE = null_mae, R2 = 0))
  }
}

# Display Final reproducible results
print(performance_lasso)

```


#### VSURF
```{r}
# Initialize results for this section
performance_vsurf <- data.frame()
selection_data_vsurf <- data.frame()
plots_importance_vsurf <- list()
plots_pred_obs_vsurf <- list()

for (response in met_vars_log) {
  # 1. Pull the standardized split for this response
  train_df <- data_splits[[response]]$train
  test_df  <- data_splits[[response]]$test
  
  # VSURF works best with matrix inputs
  x_train <- model.matrix(as.formula(paste(response, "~ .")), train_df)[,-1]
  y_train <- train_df[[response]]
  x_test  <- model.matrix(as.formula(paste(response, "~ .")), test_df)[,-1]
  y_test  <- test_df[[response]]
  
  # 2. Run VSURF (Prediction Step)
  # CRITICAL: Move seed here and set parallel = FALSE for 100% reproducibility
  set.seed(86) 
  
  # Note: VSURF is computationally heavy; without parallel it may take a moment
  vsurf_obj <- VSURF(x = x_train, y = y_train, parallel = FALSE, trace = FALSE)
  
  # varselect.pred contains the indices of variables for the prediction step
  vsurf_indices <- vsurf_obj$varselect.pred
  vsurf_vars_raw <- colnames(x_train)[vsurf_indices]
  
  if(length(vsurf_vars_raw) > 0) {
    # 3. Refit Linear Model for P-values and Standardized Coefficients
    refit_formula <- as.formula(paste(response, "~", paste(vsurf_vars_raw, collapse = " + ")))
    
    # Create temp dataframes from matrices for the lm refit
    # Use drop = FALSE to ensure it remains a matrix even if only 1 variable is selected
    refit_train <- as.data.frame(x_train[, vsurf_vars_raw, drop = FALSE])
    refit_train[[response]] <- y_train
    
    refit_test <- as.data.frame(x_test[, vsurf_vars_raw, drop = FALSE])
    # We don't necessarily need y_test in refit_test for predict(), but it's safe
    
    vsurf_refit_model <- lm(refit_formula, data = refit_train)
    
    # 4. Predictions & Performance (Testing on holdout set)
    vsurf_preds <- predict(vsurf_refit_model, newdata = refit_test)
    performance_vsurf <- rbind(performance_vsurf, 
                               get_metrics(y_test, vsurf_preds, "VSURF", response))
    
    # 5. Importance Data (Coefficients)
    coef_raw <- as.data.frame(summary(vsurf_refit_model)$coefficients)
    colnames(coef_raw) <- c("Estimate", "StdErr", "t_val", "p_val")
    
    coef_df <- coef_raw %>%
      filter(rownames(.) != "(Intercept)") %>%
      mutate(
        Predictor = clean_pred_names(rownames(.), confs),
        Outcome = response,
        Direction = ifelse(Estimate > 0, "Positive", "Negative"),
        Sig = ifelse(p_val < 0.05, "Significant", "Non-Sig")
      )
    selection_data_vsurf <- rbind(selection_data_vsurf, coef_df)
    
    # 6. Generate Plots
    # Importance Plot
    plots_importance_vsurf[[response]] <- ggplot(coef_df, aes(x = reorder(Predictor, Estimate), y = Estimate, fill = Direction)) +
      geom_bar(stat = "identity") +
      coord_flip() +
      theme_minimal() +
      labs(title = paste("VSURF Importance:", response), x = "Confounder", y = "Standardized Coefficient")
    
    # Pred vs Obs Plot
    pred_obs_df <- data.frame(Observed = y_test, Predicted = as.vector(vsurf_preds))
    plots_pred_obs_vsurf[[response]] <- ggplot(pred_obs_df, aes(x = Observed, y = Predicted)) +
      geom_point(color = "firebrick", alpha = 0.6) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
      theme_classic() +
      labs(title = paste("VSURF Pred vs Obs:", response), 
           subtitle = paste("Test R2 =", round(tail(performance_vsurf$R2, 1), 3)))
      
  } else {
    # 7. Null Model Logic: Calculate the error of the 'Mean' if no vars selected
    # This prevents the NA values in your Panel B
    null_rmse <- sqrt(mean((y_test - mean(y_train))^2))
    null_mae <- mean(abs(y_test - mean(y_train)))
  
    performance_vsurf <- rbind(performance_vsurf, 
                               data.frame(Outcome = response, Method = "VSURF", 
                                          RMSE = null_rmse, MAE = null_mae, R2 = 0))
  }
}

# Display Performance
print(performance_vsurf)
```



### Condolidate all Results

First, we combine the selection_data from all five methods into one "Master Selection" data frame.

```{r}
# Combine all selection data
master_selection <- bind_rows(
  selection_data_bw %>% mutate(Method = "Backward"),
#  selection_data_fw %>% mutate(Method = "Forward"),
  selection_data_both %>% mutate(Method = "Both"),
  selection_data_lasso %>% mutate(Method = "Lasso"),
  selection_data_vsurf %>% mutate(Method = "VSURF")
)

# Create a complete grid of all possible combinations 
# This ensures that "unselected" variables show up as gray tiles
complete_grid <- expand.grid(
  Predictor = confs,
  Outcome = met_vars_log,
  Method = c("Backward", "Forward", "Both", "Lasso", "VSURF")
)

# 1. Prepare the data with outline logic
plot_tile_data <- complete_grid %>%
  left_join(master_selection, by = c("Predictor", "Outcome", "Method")) %>%
  mutate(
    Outline_Type = case_when(
      Sig == "Significant" ~ "Solid (p < 0.05)",
      !is.na(Estimate) ~ "Dashed (Selected)",
      TRUE ~ "None"
    ),
    Method = factor(Method, levels = rev(c("Backward", "Forward", "Both", "Lasso", "VSURF"))),
    y_idx = as.numeric(Method)
  )

# Clean outcome names for matching the logic in your request
plot_tile_data <- plot_tile_data %>%
  mutate(
    DisplayOutcome = gsub("_log", "", Outcome),
    DisplayOutcome = gsub("Trig", "Triglycerides", DisplayOutcome),
    DisplayOutcome = gsub("DL", "DL-C", DisplayOutcome)
  )

# 2. Logic to Bold Selected Approaches
# We create a helper column to identify if a Method is the "Chosen" one for that Facet
plot_tile_data <- plot_tile_data %>%
  mutate(
    is_selected_approach = case_when(
      DisplayOutcome == "Glucose" & Method == "Both" ~ TRUE,         # "Both" is the Stepwise result
      DisplayOutcome == "Triglycerides" & Method == "Both" ~ TRUE,  # "Both" is the Stepwise result
      DisplayOutcome == "HDL-C" & Method == "VSURF" ~ TRUE,
      DisplayOutcome == "LDL-C" & Method == "VSURF" ~ TRUE,
      TRUE ~ FALSE
    )
  )

conf_heatmap <- ggplot(plot_tile_data, aes(x = Predictor, y = Method)) +
  geom_tile(aes(fill = Estimate), color = "white") +
  
  # Highlight the background of the selected method row slightly (optional but helpful)
  geom_rect(data = filter(plot_tile_data, is_selected_approach & Predictor == confs[1]),
            aes(xmin = -Inf, xmax = Inf, ymin = y_idx - 0.5, ymax = y_idx + 0.5),
            fill = "black", alpha = 0.2, inherit.aes = FALSE) +

  # Significant and Selected Outlines
  geom_tile(data = filter(plot_tile_data, Sig == "Significant"),
            fill = NA, color = "black", size = 0.8, linetype = "solid") +
  geom_tile(data = filter(plot_tile_data, !is.na(Estimate) & Sig != "Significant"),
            fill = NA, color = "black", size = 0.5, linetype = "dashed") +

  scale_fill_gradient2(high = "#e34a33", mid = "white", low = "#2b8cbe", 
                       midpoint = 0, na.value = "#f0f0f0",
                       name = "Coefficient Weight") +
                       
  facet_grid(DisplayOutcome ~ ., scales = "free_y", space = "free_y") +
  theme_minimal() +
  labs(title = "Confounder Selection Landscape", x = "Potential Confounders", y = "Selection Method", tag="A") +
  theme(
    plot.title = element_text(size = 20),
    axis.text.x = element_text(size=15, angle = 45, hjust = 1),
    # Note: Vectorized face labeling in theme() doesn't play perfectly with facets, 
    # so we rely on the visual indicator or manual grob editing if strictly needed.
    # However, the geom_rect above provides the visual "bolding" of the row.
    axis.text.y = element_text(size=15), 
    axis.title.x = element_text(size=15),
    axis.title.y = element_text(size=15),
    strip.text.y = element_text(size=15, angle = 0, face = "bold"),
    panel.grid = element_blank(),
     plot.tag = element_text(size = 25)
  )

# Combine all performance summaries
master_performance <- bind_rows(
  performance_bw, 
  performance_fw, 
  performance_both, 
  performance_lasso, 
  performance_vsurf
)

# Apply the same display cleaning as Panel A
master_performance <- master_performance %>%
  mutate(
    DisplayOutcome = gsub("_log", "", Outcome),
    DisplayOutcome = gsub("Trig", "Triglycerides", DisplayOutcome),
    DisplayOutcome = gsub("DL", "DL-C", DisplayOutcome),
    # Match the Y-axis factor levels exactly
    Method = factor(Method, levels = rev(c("Backward", "Forward", "Both", "Lasso", "VSURF")))
  )


# Format the performance data for the table plot
performance_table_data <- master_performance %>%
  mutate(
    # Match Panel A naming
    DisplayOutcome = gsub("_log", "", Outcome),
    DisplayOutcome = gsub("Trig", "Triglycerides", DisplayOutcome),
    DisplayOutcome = gsub("DL", "DL-C", DisplayOutcome),
    # Match Y-axis factor levels
    Method = factor(Method, levels = rev(c("Backward", "Forward", "Both", "Lasso", "VSURF"))),
    # Format numbers for display
    R2_txt = sprintf("%.3f", R2),
    RMSE_txt = sprintf("%.3f", RMSE),
    MAE_txt = sprintf("%.3f", MAE)
  ) %>%
  # Reshape to long format so we can plot RMSE, MAE, and R2 side-by-side
  pivot_longer(cols = c(R2_txt, RMSE_txt, MAE_txt), 
               names_to = "Metric", 
               values_to = "Value") %>%
  mutate(Metric = factor(Metric, levels = c("R2_txt", "RMSE_txt", "MAE_txt"),
                         labels = c("R2", "RMSE", "MAE")))

# 1. Identify the best R2 per outcome for highlighting
performance_table_data <- performance_table_data %>%
  group_by(DisplayOutcome, Metric) %>%
  mutate(
    is_best = ifelse(Metric == "R2" & Value == max(Value), "bold", "plain")
  ) %>%
  ungroup()

# 2. Build the Table Plot
conf_table <- ggplot(performance_table_data, aes(x = Metric, y = Method, label = Value)) +
  # Use aes(fontface = is_best) to highlight the top performers
  geom_text(aes(fontface = is_best), size = 5) +
  
  facet_grid(DisplayOutcome ~ ., scales = "free_y", space = "free_y") +
  
  # Move X axis labels to the top
  scale_x_discrete(position = "top") +
  
  theme_minimal() +
  labs(
    title = "Model Performance",
    x = NULL,
    y = NULL,
    tag = "B"
  ) +
  theme(
    # Repeated Y axis labels
    axis.text.y = element_text(size = 15), 
    axis.ticks.y = element_blank(),
    
    # Top-aligned X axis labels
    axis.text.x = element_text(size = 15, face = "bold"),
    
    # Facet labels
    strip.text.y = element_text(size = 15, angle = 0, face = "bold"), 
    
    panel.grid = element_blank(),
    plot.tag = element_text(size = 25),
    panel.spacing = unit(0.5, "lines"),
    
    # FIXED: Explicitly calling ggplot2::margin to avoid the 'observed' error
    plot.title = element_text(size = 20, hjust = 0.5, 
                             margin = ggplot2::margin(b = 20))
  )

library(patchwork)

# Assuming your first plot is saved as 'panel_a'
# We use widths to give the Heatmap (A) more space than the Performance (B)
conf_ab <- conf_heatmap + conf_table + 
  plot_layout(widths = c(2, 1))

```


###Final Confounders Saved
```{r}

#Confounders for specific variables 
conf_vars_HDL <- as.character(plot_tile_data$Predictor[which(!is.na(plot_tile_data$Sig) & plot_tile_data$Outcome=="HDL_log" & plot_tile_data$Method=="VSURF")])
conf_vars_LDL <- as.character(plot_tile_data$Predictor[which(!is.na(plot_tile_data$Sig) & plot_tile_data$Outcome=="LDL_log" & plot_tile_data$Method=="VSURF")])
conf_vars_Trig <- as.character(plot_tile_data$Predictor[which(!is.na(plot_tile_data$Sig) & plot_tile_data$Outcome=="Trig_log" & plot_tile_data$Method=="Both")]) 
conf_vars_Glucose <- as.character(plot_tile_data$Predictor[which(!is.na(plot_tile_data$Sig) & plot_tile_data$Outcome=="Glucose_log" & plot_tile_data$Method=="Both")]) 
conf_vars <- list(conf_vars_HDL=conf_vars_HDL,
                  conf_vars_LDL=conf_vars_LDL,
                  conf_vars_Trig=conf_vars_Trig,
                  conf_vars_Glucose=conf_vars_Glucose)

saveRDS(conf_vars, file = "../../Data/R_Data/conf_vars.RDS")



```




###PCoA and ADONIS
```{r}
# --- 1. Data Preparation & Distance Matrix ---
non_NA_rows  <- which(!is.na(rowSums(base_t1_model[, met_vars_log])))
met_ord_data <- base_t1_model[non_NA_rows, met_vars_log]
met_conf_data <- base_t1_model[non_NA_rows, confs]

met_ord_dist <- vegdist(met_ord_data, method = "euclidean", na.rm = TRUE, scale = TRUE)

# --- 2. ADONIS (Permutational MANOVA) ---
adonis_results <- data.frame(Confounder = confs, pval = NA, R2=NA)

for (r in 1:nrow(adonis_results)) {
  var_v <- adonis_results$Confounder[r]
  formula_v <- as.formula(paste0("met_ord_dist ~ ", var_v))
  adonis_v <- adonis2(formula_v, data = met_conf_data, method = "euclidean")
  adonis_results$pval[r] <- adonis_v$`Pr(>F)`[1]
  adonis_results$R2[r] <- adonis_v$R2[1]
}

adonis_results$p.signif <- add_asterisks(adonis_results$pval)
sig_confs <- adonis_results$Confounder[adonis_results$pval < 0.05]

# --- 3. PCoA Calculation ---
pcoa_res <- cmdscale(met_ord_dist, k = 2, eig = TRUE)
var_exp  <- round(pcoa_res$eig / sum(pcoa_res$eig) * 100, 1)

pcoa_df  <- data.frame(
  Axis1 = pcoa_res$points[, 1],
  Axis2 = pcoa_res$points[, 2],
  met_conf_data
)

# --- 4. Panel C: Metabolic Loading Biplot (Hollow Circles) ---
fit <- envfit(pcoa_res, met_ord_data, na.rm = TRUE)
arrows <- as.data.frame(scores(fit, display = "vectors"))
arrows$Variable <- rownames(arrows)

arrows <- arrows %>%
  mutate(
    DisplayVariable = gsub("_log", "", Variable),
    DisplayVariable = gsub("Trig", "Triglycerides", DisplayVariable),
    DisplayVariable = gsub("DL", "DL-C", DisplayVariable)
  )

conf_biplot <- ggplot() +
  # Using shape 21 with fill = NA creates the "uncovered" hollow circle effect
  geom_point(data = pcoa_df, aes(x = Axis1, y = Axis2), 
             shape = 21, color = "gray", fill = "gray", stroke = 0.5, size = 1.5) +
  
  # Vector arrows (scaled by 500 for visibility)
  geom_segment(data = arrows, 
               aes(x = 0, y = 0, xend = .6*Dim1, yend = .6*Dim2),
               arrow = arrow(length = unit(0.2, "cm")), 
               color = "black", linewidth = 0.8) +
  
  geom_text(data = arrows, 
            aes(x = .6*(Dim1+.2), y = .6*Dim2, label = DisplayVariable), 
            vjust = -0.5, hjust = 1.1, fontface = "bold", size = 5) +
  
  theme_minimal() +
  labs(
    title = "Metabolic Loading Biplot",
    x = paste0("PCoA1 (", var_exp[1], "%)"),
    y = paste0("PCoA2 (", var_exp[2], "%)"),
    tag = "C"
  ) +
  theme(
    axis.text = element_text(size = 15),
    axis.title = element_text(size = 15),
    title = element_text(size = 20),
    plot.tag = element_text(size = 25),
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA)
  )

# --- 5. Panel D: Significant ADONIS Color Grid ---
plot_list <- list()

for (i in 1:length(sig_confs)) {
  v <- sig_confs[i]
  is_num <- is.numeric(pcoa_df[[v]])
  
  p <- ggplot(pcoa_df, aes(x = Axis1, y = Axis2)) +
    geom_point(aes(color = .data[[v]]), alpha = 0.7, size = 2) +
    theme_minimal() +
    labs(
      title = paste("Colored by:", v),
      x = if (i == length(sig_confs)) paste0("PCoA1 (", var_exp[1], "%)") else NULL,
      y = paste0("PCoA2 (", var_exp[2], "%)")
    ) +
    theme(
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 15),
      title = element_text(size = 15),
      legend.title = element_text(size = 12),
      panel.grid = element_blank(),
      panel.border = element_rect(color = "grey80", fill = NA)
    )
  
  if (is_num) {
    p <- p + scale_color_viridis_c()
  } else {
    p <- p + scale_color_brewer(palette = "Set1")
  }
  
  plot_list[[v]] <- p
}

conf_adonis_plot <- wrap_plots(plot_list, ncol = 1) + 
  plot_annotation(tag_levels = list(c("D"))) & 
  theme(plot.tag = element_text(size = 25))

# --- 6. Final Layout Composition ---
conf_cd <- conf_biplot + conf_adonis_plot + 
  plot_layout(widths = c(2, 1))

print(conf_cd)

```




###Final Plots Saved
```{r}
library(ggplot2)
library(dplyr)

# 1. Create a mapping with the clean labels you want for the X-axis
winning_mapping <- data.frame(
  Outcome = c("HDL_log", "LDL_log", "Trig_log", "Glucose_log"),
  MethodLabel = c("HDL-C (VSURF)", "LDL-C (VSURF)", "Triglycerides (Stepwise)", "Glucose (Stepwise)"),
  WinningMethod = c("VSURF", "VSURF", "Both", "Both")
)

# 2. Process the data
fig1a_data <- plot_tile_data %>%
  inner_join(winning_mapping, by = c("Outcome", "Method" = "WinningMethod")) %>%
  filter(!is.na(Estimate)) %>%
  mutate(
    Point_Cat = case_when(
      Estimate > 0 & Sig == "Significant" ~ "Positive (Significant)",
      Estimate > 0 & Sig != "Significant" ~ "Positive",
      Estimate < 0 & Sig == "Significant" ~ "Negative (Significant)",
      Estimate < 0 & Sig != "Significant" ~ "Negative"
    ),
    Point_Cat = factor(Point_Cat, levels = c(
      "Positive (Significant)", "Positive", 
      "Negative (Significant)", "Negative"
    ))
  )


# 3. Create the Plot with X and Y swapped
conf_dotplot_final <- ggplot(fig1a_data, aes(x = MethodLabel, y = Predictor)) +
  # Background grid lines to help eye alignment
  geom_hline(yintercept = seq_along(levels(factor(fig1a_data$Predictor))), 
             color = "grey90", linetype = "dotted") +
  
  # The Dots
  geom_point(aes(color = Point_Cat), size = 7) +
  
  # Color Palette
  scale_color_manual(
    values = c(
      "Positive (Significant)" = "red", # Solid Red
      "Positive"       = "pink", # Light Red
      "Negative (Significant)" = "darkblue", # Solid Blue
      "Negative"       = "lightblue"  # Light Blue
    ),
    name = "Selection Type"
  ) +
  
  theme_minimal() +
  labs(
    title = "Confounder Selection",
    x = "Metabolic Outcome (Selected via Approach)",
    y = "Potential Confounders",
    tag = "A"
  ) +
theme(legend.position = "right",
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        axis.text.y = element_text(size=12),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(size = 12, angle=45, hjust=1),
        strip.text.x = element_text(size=15),
        strip.text.y = element_text(size=15),
        title = element_text(size=20),
        plot.tag = element_text(size=25))

print(conf_dotplot_final)
# Combine all panels into a single row
# Layout: [Heatmap A] [Table B] [Biplot C] [ADONIS D]
conf_final_row <- conf_heatmap + conf_table + conf_biplot + conf_adonis_plot + 
  plot_layout(
    nrow = 1, 
    widths = c(2, 1, 3, 1) # Adjusts relative space for wider panels A & C
  )

# Print the final figure
print(conf_final_row)
#The plot is saved

saveRDS(conf_final_row, file = "../../Figures/Panels/ConfFig.RDS")

saveRDS(conf_dotplot_final, file = "../../Figures/Panels/FIG1A.RDS")


```

